Remark on the hypothetical judgment

### 2015-08-01 paper.tex
explain unary logical relations

\documentclass[11pt]{amsart}
\usepackage{setspace}
\usepackage{amssymb}
\usepackage{proof}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[thm]{Definition}
\newtheorem{example}[thm]{Example}
%\newtheorem{note}[thm]{Note}
\theoremstyle{remark}
\newtheorem{remark}[thm]{Remark}
\numberwithin{equation}{section}

@@

\newcommand\HypJ[2]{#1\ (#2)}
\newcommand\GenJ[2]{\vert_{#1}\; #2}
\newcommand\JJ{\mathcal{J}}
\newcommand\IsVer[1]{\InputMode{#1}\ \mathit{ver}}
\newcommand\IsTrue[1]{\InputMode{#1}\ \mathit{true}}
\newcommand\IsProp[1]{\InputMode{#1}\ \mathit{prop}}
\newcommand\Member[2]{\InputMode{#1}\in \InputMode{#2}}
\newcommand\Seq[2]{\InputMode{#1}\vdash #2}
\newcommand\Nat{\mathbb{N}}

@@

In 1956, Heyting propounded his version of what has come to be known as the
\emph{Brouwer-Heyting-Kolmogorov} interpretation of intuitionistic logic, by
explaining the assertion conditions of the propositions. Note that where Heyting
says ``assert a proposition'', in light of Martin-L\"of's clarification, we must
understand ``assert \emph{the truth of} a proposition''. Heyting's explanation
of the assertion conditions for the truth of implication were as follows:
\begin{quote}
  The implication $p\to q$ may be asserted if and only if we possess a
  construction $r$, which, joined to any construction proving $p$ (supposing the
  latter be effected), would automatically effect a construction of $q$. \cite{HeytingA:int}
\end{quote}

@@

\begin{quote}
$\IsTrue{P\to Q}$ may be asserted if and only if we
possess a construction $r$, which, joined to any demonstration of $\IsTrue{P}$,
would automatically effect a demonstration of $\IsTrue{Q}$.
\end{quote}

@@

\begin{quote}
  $\IsTrue{P\to Q}$ may be asserted if and only if we may assert $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
\end{quote}

In fact, if we make this transformation, we shall have arrived at something very
similar to Martin-L\"of's definition of propositional implication. This
inference is, at least, valid with respect to Martin-L\"of's definition, but it
is merely an extensional specification for the meaning of the judgment: it
expresses the material equivalence of the assertions $\IsTrue{P\to Q}$ and
$\HypJ{\IsTrue{Q}}{\IsTrue{P}}$, but it does not contain an actual
\emph{explanation} of $\IsTrue{P\to Q}$, which would need to be in the form ``To
know $\IsTrue{P\to Q}$ is to know...''.

@@

\begin{quote}
  To know $\IsVer{P\to Q}$ is to know $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
\end{quote}

Then, to know $\IsTrue{P\to Q}$ is to have a means of verifying $P\to Q$, that
is, to have a plan to experience $\IsVer{P\to Q}$.

@@

If we return to Heyting's original definition of the assertion $\IsTrue{P\to
Q}$, it is clear that he would have to accept any mathematical means of
transforming the construction of $\IsTrue{P}$ into a construction of
$\IsTrue{Q}$, since the construction $r$ in his definition is not constrained by
any uniformity condition: it is merely any effective operation which, when
adjoined with a construction of the premise, effects a construction of the
conclusion; and moreover, $r$ may proceed by virtue of any of the transcendental
observations which are possible by virtue of the truth of $P$ having been
experienced by the subject.

@@

%%% TODO:
%%% - realizability as unary logical relations
%%% - (nuprl) type theory as binary logical relations
%%% - what do the propositions give us over the judgments? (witnesses, extensional identity)
%%% - harper's presentation of admissibility & derivability in pfpl
%%% - noam zeilberger's use of non-uniform hypothetical judgment

>>>>

\documentclass[11pt]{amsart}
\usepackage{setspace}
\usepackage{amssymb,stmaryrd}
\usepackage{proof}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[thm]{Definition}
\newtheorem{example}[thm]{Example}
%\newtheorem{note}[thm]{Note}
\theoremstyle{remark}
\newtheorem{remark}[thm]{Remark}
\numberwithin{equation}{section}

@@

\newcommand\HypJ[2]{#1\ (#2)}
\newcommand\GenJ[2]{\vert_{#1}\; #2}
\newcommand\JJ{\mathcal{J}}
\newcommand\IsVer[1]{\InputMode{#1}\ \mathit{ver}}
\newcommand\IsTrue[1]{\InputMode{#1}\ \mathit{true}}
\newcommand\IsProp[1]{\InputMode{#1}\ \mathit{prop}}
\newcommand\IsSet[1]{\InputMode{#1}\ \mathit{set}}
\newcommand\Member[2]{\InputMode{#1}\in \InputMode{#2}}
\newcommand\Seq[2]{\InputMode{#1}\vdash #2}
\newcommand\Eval[2]{\InputMode{#1}\Rightarrow \OutputMode{#2}}
\newcommand\Nat{\mathbb{N}}

\newcommand\SEM[1]{\llbracket#1 \rrbracket}
\newcommand\VAL[1]{\mathcal{V}\SEM{#1}}
\newcommand\EXP[1]{\mathcal{E}\SEM{#1}}

\newcommand\True{\top}
\newcommand\False{\bot}
\newcommand\Imp[2]{#1\supset #2}
\newcommand\Conj[2]{#1\land #2}
\newcommand\It{\star}
\newcommand\Lam[2]{(\lambda #1) #2}
\newcommand\Pair[2]{\langle #1, #2 \rangle}
\newcommand\Fst{\mathsf{fst}}
\newcommand\Ap{\mathsf{ap}}
\newcommand\IsDefined[1]{\InputMode{#1}\ \mathit{defined}}

@@

In 1956, Heyting propounded his version of what has come to be known as the
\emph{Brouwer-Heyting-Kolmogorov} interpretation of intuitionistic logic, by
explaining the assertion conditions of the propositions. Note that where Heyting
says ``assert a proposition'', in light of Martin-L\"of's clarification, we must
understand ``assert \emph{the truth of} a proposition''. Heyting's explanation
of the assertion conditions for the truth of implication were as follows:
\begin{quote}
  The implication $\Imp{p}{q}$ may be asserted if and only if we possess a
  construction $r$, which, joined to any construction proving $p$ (supposing the
  latter be effected), would automatically effect a construction of $q$. \cite{HeytingA:int}
\end{quote}

@@

\begin{quote}
$\IsTrue{\Imp{P}{Q}}$ may be asserted if and only if we
possess a construction $r$, which, joined to any demonstration of $\IsTrue{P}$,
would automatically effect a demonstration of $\IsTrue{Q}$.
\end{quote}

@@

\begin{quote}
  $\IsTrue{\Imp{P}{Q}}$ may be asserted if and only if we may assert $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
\end{quote}

In fact, if we make this transformation, we shall have arrived at something very
similar to Martin-L\"of's definition of propositional implication. This
inference is, at least, valid with respect to Martin-L\"of's definition, but it
is merely an extensional specification for the meaning of the judgment: it
expresses the material equivalence of the assertions $\IsTrue{Imp{P}{Q}}$ and
$\HypJ{\IsTrue{Q}}{\IsTrue{P}}$, but it does not contain an actual
\emph{explanation} of $\IsTrue{Imp{P}{Q}}$, which would need to be in the form ``To
know $\IsTrue{Imp{P}{Q}}$ is to know...''.

@@

\begin{quote}
  To know $\IsVer{Imp{P}{Q}}$ is to know $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
\end{quote}

Then, to know $\IsTrue{Imp{P}{Q}}$ is to have a means of verifying $Imp{P}{Q}$, that
is, to have a plan to experience $\IsVer{Imp{P}{Q}}$.

@@

If we return to Heyting's original definition of the assertion
$\IsTrue{\Imp{P}{Q}}$, it is clear that he would have to accept any mathematical
means of transforming the construction of $\IsTrue{P}$ into a construction of
$\IsTrue{Q}$, since the construction $r$ in his definition is not constrained by
any uniformity condition: it is merely any effective operation which, when
adjoined with a construction of the premise, effects a construction of the
conclusion; and moreover, $r$ may proceed by virtue of any of the transcendental
observations which are possible by virtue of the truth of $P$ having been
experienced by the subject.

@@

\subsection{Realizability Models as Unary Logical Relations}

In fact, by replacing the abstract/synthetic explanations of the propositions in
terms of their verification acts, we can recast them as giving rise to a theory
of sets or types by explaining them in terms of their verification objects (i.e.
their canonical witnesses).

A realizability model in this simple sense amounts to interpreting the
propositions into unary logical relations. Defining a proposition, then, amounts
to defining the unary relation $\VAL{P}$ (which is the set of verification
objects of the proposition $P$); then, a separate logical relation $\EXP{P}$ is
defined uniformly over all propositions $P$ by appealing to the computation
$\Eval{M}{N}$ of witnesses to canonical form.
\begin{align*}
  \VAL{\True} &\equiv \{\It\}\\
  \VAL{\False} &\equiv \{\}\\
  \VAL{\Imp{P}{Q}} &\equiv 
    \{ \Lam{x}{E}
    \mid \GenJ{x}{\HypJ{\Member{E}{\EXP{Q}}}{\Member{x}{\EXP{P}}}}
    \}\\
  \VAL{\Conj{P}{Q}} &\equiv
    \{ \Pair{M}{N}
    \mid \Member{M}{\EXP{P}}, \Member{N}{\EXP{Q}}
    \}\\
  \EXP{P} &\equiv
    \{ M
    \mid \Eval{M}{M'}, \Member{M'}{\VAL{P}}
    \}
\end{align*}

Then, the judgments of the logical theory are explained by appealing to the
logical relations:

\begin{gather*}
  \infer={
    \IsProp{P}
  }{
    \IsDefined{\VAL{P}}
  } \qquad
  \infer{
    \IsVer{P}
  }{
    \Member{M}{\VAL{P}}
  }\qquad
  \infer{
    \IsTrue{P}
  }{
    \Member{M}{\EXP{P}}
  }
\end{gather*}

The judgments of a basic theory of sets may also be explained using the logical
relations:
\begin{gather*}
  \infer={
    \IsSet{A}
  }{
    \IsDefined{\VAL{A}}
  } \qquad
  \infer={
    \Member{M}{A}
  }{
    \Member{M}{\EXP{A}}
  }
\end{gather*}

The material interpretation of the hypothetical judgment is crucial in the
realizability model; this is because only the verification objects (i.e.
canonical witnesses) are given. All the non-canonical witnesses are explained
via computation, and the use of logical consequence instead of material
consequence in the explanation of $\VAL{\Imp{P}{Q}}$ would have been disastrous.

For instance, it should be the case that $\Lam{x}{\Pair{\It}{\It}}$ is a witness
of $\Imp{\False}{\True}$. To see if this is the case in the
model, let us translate this into a concrete statement:
\begin{gather}
  \Member{\Lam{x}{\Pair{\It}{\It}}}{\EXP{\Imp{\False}{\True}}}\\
  \Eval{\Lam{x}{\Pair{\It}{\It}}}{\Lam{x}{\Pair{\It}{\It}}}, \quad
    \Member{\Lam{x}{\Pair{\It}{\It}}}{\VAL{\Imp{\False}{\True}}}\\
  \GenJ{x}{
    \HypJ{\Member{\Pair{\It}{\It}}{\EXP{\True}}}{\Member{x}{\EXP{\False}}}
  }\\
  \GenJ{x}{
    \HypJ{\Member{\Pair{\It}{\It}}{\EXP{\True}}}{\Eval{x}{M}, \Member{M}{\VAL{\False}}}
  }
\end{gather}

And at this time, we may discharge the entire hypothetical judgment, since we
know that the unary relation $\VAL{\False}$ is empty, and so there can be no
such $M$.

If we did not have the material consequence at our disposal, then this statement
would not have been valid, unless we were to eschew the verificationist meaning
explanation and also add ``use'' rules (i.e. direct elimination) for each
proposition in addition to the verification rules.

%%% TODO:
%%% - (nuprl) type theory as binary logical relations
%%% - what do the propositions give us over the judgments? (witnesses, extensional identity)
%%% - harper's presentation of admissibility & derivability in pfpl
%%% - noam zeilberger's use of non-uniform hypothetical judgment
